---
title: "3. MacroAnatomic Diversity of Cell Types"
author: "Paula Restrepo"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: yes
    keep_md: yes
    highlight: kate
---

```{r setup, include=FALSE}
# load Libraries
library(tidyverse)
library(Seurat)
library(Matrix)

library(edgeR)
library(limma)
library(variancePartition)
library(dreamlet)
library(crumblr)
library(SingleCellExperiment)

library(vegan)
library(microbiome)

library(circlize)
library(ComplexHeatmap)
library(RColorBrewer)
library(paletteer)
library(ggrepel)
library(ggbeeswarm)
library(ggpubr)
library(patchwork)
library(svglite)

# Plotting Parameters

## ggplot themes
theme <- theme(
  legend.position = 'right',
  text = element_text(color = 'black', size = 10),
  title = element_text(face = 'bold', size = 12, color = 'black', hjust = 0.5),
  strip.background = element_rect(fill = 'black', color = 'black'),
  strip.text = element_text(color = 'white', face = 'bold', size = 12),
  plot.title = element_text(size = 14, color = 'black', face = 'bold', hjust = 0.5),
  panel.background = element_rect(color = 'black', fill = NA),
  plot.background = element_blank(), 
  legend.background = element_blank(), 
  legend.key = element_blank()
  )

spatial_theme <- theme(
  rect = element_blank(),
  line = element_blank(),
  text = element_text(color = 'black', size = 10),
  title = element_text(face = 'bold', size = 12, color = 'black', hjust = 0.5),
  strip.background = element_rect(fill = 'black', color = 'black'),
  strip.text = element_text(color = 'white', face = 'bold', size = 12),
  plot.title = element_text(size = 14, color = 'black', face = 'bold', hjust = 0.5),
  axis.text = element_blank()
  )
```

## Load Data

First load the seurat object. We remove doublets, unannotated cells, and cells that fall outside the tissue per manual compartment annotation.

```{r load-seurat-object}
obj <- qs::qread("data/merfish/BAYSOR/seurat_objects/ns-atlas.merfish_baysor.scanvi_integrated.cellcharter.annotated.seurat_object.QS")

obj <- subset(obj, 
              subset = 
                cell_type.detailed != "Doublet" & 
                cell_type.detailed != "Unknown" & 
                tissue_compartment != "outside tissue")

# remove forearm and chest since they don't have enough donors (min 3+ unique donors per site for stats)
obj <- subset(obj, subset = anatomic_site != "Forearm" & anatomic_site != "Chest")
invisible(gc())

meta <- obj@meta.data

cell_types <- levels(meta$cell_type.detailed)
cell_types <- cell_types[! cell_types %in% c("Doublet", "Unknown")]
names(cell_types) <- make.names(cell_types)
names(cell_types)[42] <- "Naive.T"
```


```{r prepare-pseudobulk}
sce <- SingleCellExperiment(assays = list(counts = obj[['RNA']]$counts), colData = obj@meta.data)

pb <- aggregateToPseudoBulk(sce,
                            assay = "counts",
                            cluster_id = "cell_type.detailed",
                            sample_id = "sample_barcode",
                            verbose = FALSE)
assayNames(pb)
pb$sample_barcode <- colnames(pb)
counts <- cellCounts(pb)
rm(sce); gc()
```

```{r load-compartment-areas}
tissue_areas <- read.csv('data/merfish/BAYSOR/metadata/tissue_compartment_areas.csv')
tissue_areas$sample_barcode <- substr(tissue_areas$sample_compartment, 1, nchar(tissue_areas$sample_compartment)-4)
tissue_areas <- tissue_areas[, c("sample_barcode", 'total_area_mm2',"compartment.short",'compartment_area_mm2')]

tissue_areas <- tissue_areas %>% 
  pivot_wider(names_from = 'compartment.short', values_from = 'compartment_area_mm2', names_prefix = "AREA_", values_fill = 0)
colnames(tissue_areas) <- c("sample_barcode", "total_area_mm2", "dermis_area_mm2", 'epidermis_area_mm2', 'subcutis_area_mm2', 'stratcorn_area_mm2')
head(tissue_areas)
```
## Basic Cell Density and Diversity Across Sites

```{r calc-density-and-diversity}
info <- data.frame(colData(pb))
info$anatomic_site <- make.names(info$anatomic_site)
info$total_area_mm2 <- NULL
info <- left_join(info, tissue_areas)
info <- data.frame(info)
rownames(info) <- info$sample_barcode
density <- (counts / info$total_area_mm2) / 100 # tx per 100um2
info$shannon_diversity <- diversity(t(density), index = 'shannon')[,1]

total.cells <- rowSums(counts)
info$total.cells <- total.cells
info$cell_density <- (info$total.cells / info$total_area_mm2) / 100 # tx per 100um2
info$site_short <- reorder(info$site_short, info$shannon_diversity, median)
```

Density correlates with cell diversity

```{r density_vs_diversity_scatterplot}
ggplot(info) + 
  aes(x = cell_density, y = shannon_diversity) + 
  geom_point(show.legend=FALSE) + theme_bw()  + 
  geom_text_repel(aes(label=sample_barcode), size=3, min.segment.length = 0, point.padding = 0) +
  geom_smooth(method='lm', color='red') + scale_x_log10()
```


```{r FIGURE_3A_DensityAndDiversity_PerAnatomicSite_Boxplot, fig.height=5, fig.width=3.5}
# average across biological replicates
info2 <- info %>% 
  group_by(donor_id, site_short) %>% 
  summarise(shannon_diversity=mean(shannon_diversity), 
            cell_density=mean(cell_density)) %>% 
  ungroup()
global_info2 <- info2
global_info <- info

site_order <- levels(reorder(info2$site_short, info2$shannon_diversity, median))
div <- ggplot(info2) + 
  aes(x = reorder(site_short, shannon_diversity, median), 
      y = shannon_diversity, 
      color = site_short) + 
  geom_quasirandom(show.legend=FALSE)+
  geom_boxplot(alpha=0, aes(color=NULL), show.legend=FALSE) + 
  theme_classic() + rotate_x_text() +
  scale_color_paletteer_d("Polychrome::light") +
  labs(x = NULL, y = "Shannon Diversity Index") + theme

den <- ggplot(info2) + 
  aes( x = reorder(site_short, shannon_diversity, median), 
       y = cell_density, color = site_short) + 
  geom_quasirandom(show.legend=FALSE)+
  geom_boxplot(alpha=0, aes(color=NULL), width=0.8, show.legend=FALSE) + 
  theme_classic() + rotate_x_text() +
  scale_color_paletteer_d("Polychrome::light") +
  labs(x = NULL, y = "Cell Density") + theme

FIG3A <- (div / den) + plot_layout(axes='collect_x')

FIG3A
```



```{r SAVE_FIGURE3A, include=FALSE}
svglite::svglite("figures/merfish/BAYSOR/shannon_diversity/FIGURE3A_FINAL.tissue_area_normalized.no_compartments.shannon_index_and_density_panel.svg", height=5, width=3.5)
FIG3A
dev.off()
```

```{r FIGURE3B-make-dots-for-12-donor-graphicpanel}
FIG3B <- ggplot(info2) + 
  aes(x=donor_id, y=site_short, size=cell_density, color=shannon_diversity) + 
  geom_point() + theme_classic()+
  viridis::scale_color_viridis(option = 'A') +
  rotate_x_text() + theme +
  labs(x="Donor ID", y='Anatomic Site')

FIG3B
```


```{r SAVE_FIGURE3B, include=FALSE}
svglite::svglite("figures/merfish/BAYSOR/shannon_diversity/FIGURE3B.median_diversity_density_dots_per_anatomic_site_per_donor.svg")
FIG3B
dev.off()
```

## Crumblr: Differential Abundance

```{r make-crumblr-object}
## crumblr must be done with COUNTS not densities (per convo with Gabe Hoffman)
cobj <- crumblr(counts)
```


### Crumblr PCA

```{r calculate-pca-with-crumblr}
pca <- prcomp(t(standardize(cobj)), scale. = TRUE, center=TRUE)
variance_proportions = round(summary(pca)$importance[2,]*100, digits=2)

# merge with metadata
df_pca <- merge(pca$x, info, by = "row.names")

umap <- uwot::umap(t(standardize(cobj)))
colnames(umap) <- c("UMAP_1", 'UMAP_2')
df_umap <- merge(umap, info, by = "row.names")
```



```{r make-crumblr-pca-plots, fig.height=6.5, fig.width=20}
anatomic_site.pca_plot <- ggplot(df_pca) +
  aes(x = PC1, y = PC2, 
      color = site_short) +
  geom_point(size = 5) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) +
  scale_color_paletteer_d("Polychrome::light")

batch.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2, color = collection_source, shape = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_shape_manual(values = c(4,1)) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 

diversity.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2, 
      fill = shannon_diversity, 
      size = cell_density) +
  geom_point(shape=21) +
  viridis::scale_fill_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 
```

```{r crumbr_pca_panel, fig.height=6, fig.width=20}
(anatomic_site.pca_plot | batch.pca_plot |  diversity.pca_plot) & theme
```


```{r}
ggplot(df_umap) +
  aes(x = UMAP_1, y = UMAP_2, 
      color = site_short) +
  geom_point(size = 5) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_color_paletteer_d("Polychrome::light")


ggplot(df_umap) +
  aes(x = UMAP_1, y = UMAP_2, 
      color = donor_id) +
  geom_point(size = 5) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_color_paletteer_d("pals::polychrome")

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, color = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) 

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, color = collection_source) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) 

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, color = batch, shape = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() + 
  theme(aspect.ratio = 1) +
  scale_shape_manual(values = c(4,1)) +
  scale_color_paletteer_d('pals::polychrome')

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, 
      fill = shannon_diversity, 
      size = cell_density) +
  geom_point(shape=21) +
  viridis::scale_fill_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) 
```


```{r run-variance-partition}
form <-  ~ donor_age + 
  total_area_mm2 + 
  epidermis_area_mm2 + 
  dermis_area_mm2 +
  subcutis_area_mm2 +
  stratcorn_area_mm2 +
  (1|anatomic_site) + 
  (1|collection_type) + 
  (1|donor_sex) + 
  (1|donor_id) + 
  (1|batch)

info$anatomic_site <- gsub("\\.", "_", gsub(" ", "",info$site_short))

vp <- fitExtractVarPartModel(cobj, form, info)
```


```{r full_variancepartition_plot, fig.height=4, fig.width=10}
vpdf <- data.frame(vp)
cols <- colnames(vpdf)
vpdf$cell_type.detailed <- factor(rownames(vpdf), levels = cell_types)
vpdf <- vpdf %>% pivot_longer(cols = cols, names_to='variable', values_to='value')
vpdf$variable <- factor(vpdf$variable, levels = cols)
color_map <- c(colorway::varibow(n=length(cols)-1), "gray80")
names(color_map) <- cols

vplot <- ggplot(vpdf) + 
  aes(x = factor(cell_type.detailed, 
                 levels = rev(row.names(data.frame(vp))[order(data.frame(vp)$Residuals)]),
                 labels = cell_types), 
      y=value, fill = variable) + 
  geom_bar(stat='identity', color='black') + 
  scale_fill_manual(values = color_map, name = NULL) + 
  theme_classic() + theme + rotate_x_text() + 
  labs(y = "Variance Explained (%)", x = "Cell Type") + 
  scale_y_continuous(labels = scales::percent)

vplot
```

Based on this, we can correct for dermis, epidermis, and subcutis area, age, sex, donor age, donor, and batch. 

```{r run-crumbr}
form <- ~ +
  donor_age + 
  epidermis_area_mm2 + 
  dermis_area_mm2 +
  subcutis_area_mm2 +
  (1|anatomic_site) +
  (1|donor_sex) + 
  (1|donor_id) + 
  (1|batch)
vp <- fitExtractVarPartModel(cobj, form, info)


form <- ~ 0 +
  donor_age + 
  epidermis_area_mm2 + 
  dermis_area_mm2 +
  subcutis_area_mm2 +
  anatomic_site +
  (1|donor_sex) + 
  (1|donor_id) + 
  (1|batch)

sites <- unique(as.character(info$anatomic_site))

contrast_list <- lapply(sites, function(site){

  side1 <- paste0("anatomic_site", site)
  side2 <- sites[sites != site]
  len <- length(side2)

  side2 <- paste0("(", paste(paste0("anatomic_site", side2), collapse=" + "), ") / ", len)

  contr <- paste0(side1, " - ", side2)

})
names(contrast_list) <- sites

L <- makeContrastsDream(form, info, contrasts = contrast_list)
fit <- dream(cobj, form, info, L)
fit <- eBayes(fit, robust = TRUE)

resdf <- lapply(1:length(sites), function(x){
  df <- data.frame(topTable(fit, number = Inf, coef = x, confint = TRUE))
  df$anatomic_site <- sites[x]
  df$cell_type.detailed <- factor(rownames(df), levels = cell_types, labels=names(cell_types))
  return(df)
}) %>% bind_rows()

table(resdf$adj.P.Val < 0.05)
table(resdf$anatomic_site, resdf$adj.P.Val < 0.05)
```






```{r convert-crumblr-results-to-matrix}
lfc.mat <- resdf %>%
  dplyr::select(anatomic_site, cell_type.detailed, logFC) %>%
  pivot_wider(names_from = "cell_type.detailed", values_from = "logFC") %>%
  data.frame()

rownames(lfc.mat) <- lfc.mat$anatomic_site
lfc.mat$anatomic_site <- NULL
lfc.mat <- lfc.mat[, names(cell_types)]
colnames(lfc.mat) <- cell_types

p.mat <- resdf %>% 
  mutate(padj.star = gtools::stars.pval(adj.P.Val)) %>% # gsub("\\.", "", gtools::stars.pval(adj.P.Val))) %>%
  dplyr::select(anatomic_site, cell_type.detailed, padj.star) %>%
  pivot_wider(names_from = "cell_type.detailed", values_from = "padj.star") %>%
  data.frame()
rownames(p.mat) <- p.mat$anatomic_site
p.mat$anatomic_site <- NULL
p.mat <- p.mat[, names(cell_types)]
colnames(p.mat) <- cell_types
```


```{r cluster-celltypes-by-diff-abd-pattern}
set.seed(123)
site.hist <- hclust(dist(as.matrix(lfc.mat)))
site.dend <- as.dendrogram(site.hist)

site_km_list <- lapply(2:8, function(k){
  site.km <- kmeans(lfc.mat, centers = k, iter.max = 1000, nstart = 100)
})
names(site_km_list) <- paste0("K", 2:8)
site_km_lab_list <- lapply(site_km_list, function(site.km){factor(paste0("C", site.km$cluster))})
```


```{r}
library(cluster) 
silhouette_score <- function(km, df){
  ss <- silhouette(km$cluster, dist(as.matrix(df)))
  mean(ss[, 3])
}
sil_scores <- sapply(site_km_list, function(km){
  silhouette_score(km, df = lfc.mat)
})

sdf <- data.frame(K=names(site_km_list), silhouette_score = sil_scores)
ggplot(sdf) + aes(x=K, y=silhouette_score) + geom_bar(stat='identity') + theme_bw()
```


```{r cluster-celltypes-by-diff-abd-pattern}
ct.hist <- hclust(dist(as.matrix(t(lfc.mat))))
ct.dend <- as.dendrogram(ct.hist)


ct_km_list <- lapply(2:10, function(k){
  site.km <- kmeans(t(lfc.mat), centers = k, iter.max = 1000, nstart = 100)
})
names(ct_km_list) <- paste0("K", 2:10)
ct_km_lab_list <- lapply(ct_km_list, function(ct.km){factor(paste0("C", ct.km$cluster))})
```

```{r}
sil_scores <- sapply(ct_km_list, function(km){
  silhouette_score(km, df = t(lfc.mat))
})

sdf <- data.frame(K=names(ct_km_list), silhouette_score = sil_scores)
ggplot(sdf) + aes(x=K, y=silhouette_score) + geom_bar(stat='identity') + theme_bw()
```




```{r fig.height=10, fig.width=14}
vpdf <- data.frame(vp)
cols <- colnames(vpdf)
vpdf$cell_type.detailed <- factor(rownames(vpdf), levels = cell_types, labels=names(cell_types))
vpdf <- vpdf %>% 
  pivot_longer(cols = cols, names_to='variable', values_to='value') %>%
  mutate(variable=factor(variable, levels=cols))

color_map <- c(colorway::varibow(n=length(cols)-1), "gray80")
names(color_map) <- cols

bp = HeatmapAnnotation(var=anno_barplot(data.frame(vp), bar_width=1, gp = gpar(fill=color_map)),
                       annotation_name_side = 'left',annotation_height = unit(2, 'cm'),
                       annotation_legend_param = Legend(at=names(color_map), legend_gp = gpar(fill = color_map))
    )

lgd = Legend(at=names(color_map), legend_gp = gpar(fill = color_map))


hm = Heatmap(
  as.matrix(lfc.mat), 
  name = "logFC", 
  row_split = site_km_lab_list[["K7"]],
  column_split = ct_km_lab_list[['K5']],
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%s", p.mat[i, j]), x, y, gp = gpar(fontsize = 10))
    },
  width = ncol( as.matrix(lfc.mat))*unit(5, "mm"), 
  height = nrow( as.matrix(lfc.mat))*unit(5, "mm"),
  top_annotation = bp,
  row_names_side = "left",
  column_dend_side = 'bottom',
  column_names_side = "bottom",
  column_title_side = "bottom"
        )

draw(hm, annotation_legend_list = list(lgd), merge_legend=TRUE)
```


```{r}
svglite::svglite("figures/merfish/BAYSOR/differential_abundance/Fig3D_FINAL.all_samples.density_adj.crumblr.comp_areas_as_terms.svg", height=10, width=14)
draw(hm, annotation_legend_list = list(lgd), merge_legend=TRUE)
dev.off()
```

## Adjusted Crumblr PCA

```{r}
form <- ~ 0 +
  donor_age + 
  epidermis_area_mm2 + 
  dermis_area_mm2 +
  subcutis_area_mm2 +
  (1|donor_sex) + 
  (1|donor_id) + 
  (1|batch)
fit <- dream(cobj, form, info)
adj_propmat <- variancePartition::residuals.MArrayLM2(fit, cobj$E)

pca <- prcomp(t(adj_propmat), scale. = TRUE, center=TRUE)
variance_proportions = round(summary(pca)$importance[2,]*100, digits=2)

# merge with metadata
df_pca <- merge(pca$x, info, by = "row.names")
set.seed(12345)
umap <- uwot::umap(t(adj_propmat), min_dist=0.1, n_neighbors = 10, seed = 12345)
colnames(umap) <- c("UMAP_1", 'UMAP_2')
df_umap <- merge(umap, info, by = "row.names")

adj_propmat2 <- t(adj_propmat)[, cell_types]
```




```{r make-adjusted-crumblr-pca-plots, fig.height=6.5, fig.width=20}
anatomic_site.pca_plot <- ggplot(df_pca) +
  aes(x = PC1, y = PC2, 
      color = site_short) +
  geom_point(size = 5) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) +
  scale_color_paletteer_d("Polychrome::light")

batch.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2, color = collection_source, shape = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_shape_manual(values = c(4,1)) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 

diversity.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2, 
      fill = shannon_diversity, 
      size = cell_density) +
  geom_point(shape=21) +
  viridis::scale_fill_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 
```

```{r adjusted_crumbr_pca_panel, fig.height=6, fig.width=20}
(anatomic_site.pca_plot | batch.pca_plot |  diversity.pca_plot) & theme
```


```{r adjusted-crumblr-umap-plots}
ggplot(df_umap) +
  aes(x = UMAP_1, y = UMAP_2, 
      color = site_short) +
  geom_point(size = 3) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_color_paletteer_d("Polychrome::light") +
  theme


ggplot(df_umap) +
  aes(x = UMAP_1, y = UMAP_2, 
      color = donor_id) +
  geom_point(size = 5) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_color_paletteer_d("pals::polychrome")

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, color = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) 

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, color = collection_source) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) 

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, color = batch, shape = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() + 
  theme(aspect.ratio = 1) +
  scale_shape_manual(values = c(4,1)) +
  scale_color_paletteer_d('pals::polychrome')

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, 
      fill = shannon_diversity, 
      size = cell_density) +
  geom_point(shape=21) +
  viridis::scale_fill_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) + theme

ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, 
      color=dermis_area_mm2) +
  geom_point() +
  viridis::scale_color_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) + theme
ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, 
      color=epidermis_area_mm2) +
  geom_point() +
  viridis::scale_color_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) + theme
ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, 
      color=subcutis_area_mm2) +
  geom_point() +
  viridis::scale_color_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) + theme
```

```{r}
site <- ggplot(df_umap) +
  aes(x = UMAP_1, y = UMAP_2, 
      color = site_short) +
  geom_point(size = 3) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_color_paletteer_d("Polychrome::light") +
  theme
divden <- ggplot(df_umap) +
  aes(UMAP_1, UMAP_2, 
      fill = shannon_diversity, 
      size = cell_density) +
  geom_point(shape=21) +
  viridis::scale_fill_viridis(option="A") +
  theme_classic() +
  theme(aspect.ratio = 1) + theme

svglite::svglite("figures/merfish/BAYSOR/differential_abundance/FigS3AB_FINAL.all_samples.adj_clr.umap.svg", height=5, width=10)
site | divden
dev.off()
```
## Adjusted CLR boxplots

```{r}
adj_clr_long <- cbind(info, adj_propmat2) %>% 
  pivot_longer(cols=colnames(adj_propmat2), names_to = 'cell_type.detailed', values_to='Adj.CLR') %>%
  mutate(cell_type.detailed = factor(cell_type.detailed, levels=cell_types))
```

```{r fig.height=7.5, fig.width=16}
clr_boxplots <- ggplot(adj_clr_long) +
  aes(x = site_short, y = Adj.CLR, color=site_short) + 
  geom_quasirandom() +
  geom_boxplot(aes(color=NULL), alpha=0, outliers = FALSE, show.legend=FALSE) +
  theme_bw() + theme +
  facet_wrap(~ cell_type.detailed, scales='free_y',nrow=5) +
  rotate_x_text() +
  scale_color_paletteer_d('Polychrome::light', name=NULL) +
  labs(x=NULL, y = "Adj. CLR") 

svglite::svglite("figures/merfish/BAYSOR/differential_abundance/FigS4D_AdjCLR_Boxplots.svg", height=7.5, width = 16)
clr_boxplots
dev.off()

clr_boxplots
```



## Correlation with Epidermal Thickness



```{r}
thicknessdf <- readRDS("data/histology/results/qupath_summary_wide.RDS")  %>% 
  mutate(sample_barcode = paste(donor_id, "SKIN_NS", site_id, tissue_replicate, sep='_')) %>%
  dplyr::select(-anatomic_site)
```

```{r fig.height=8, fig.width=4}
thicknessdf <- readRDS("data/histology/results/qupath_summary_wide.RDS")  %>% 
  mutate(sample_barcode = paste(donor_id, "SKIN_NS", site_id, tissue_replicate, sep='_')) %>%
  dplyr::select(-anatomic_site)

thicknesslong <- thicknessdf %>% 
  pivot_longer(cols=c("stratum.corneum", 'epidermis', 'total_thickness'), values_to = 'thickness', names_to = 'var') %>%
  inner_join(info[, c("sample_barcode", 'anatomic_site')])
thicknesslong
thicknesslong$anatomic_site <- factor(thicknesslong$anatomic_site)
thicknesslong$anatomic_site <- reorder(thicknesslong$anatomic_site, thicknesslong$thickness, max, na.rm=TRUE)
thickness_plot <- ggplot(thicknesslong) + 
  aes(x = anatomic_site, y = thickness, color=anatomic_site) + 
  geom_quasirandom(show.legend=FALSE) +
  ## geom_line(aes(x = as.numeric(anatomic_site), group = donor_id, color=NULL), alpha=0.1, show.legend = FALSE) +
  geom_boxplot(alpha=0, aes(color=NULL), width=0.8, show.legend=FALSE) + 
  theme_classic() + rotate_x_text() +
  scale_y_log10() +  facet_wrap(~var, ncol=1, scales='free_y') + 
  scale_color_paletteer_d("Polychrome::light") +
  labs(x = NULL, y = "Thickness (μm)") + theme


svglite::svglite("figures/merfish/BAYSOR/differential_abundance/FS4C_EpidermalThicknessAcrossSites.svg", height=6, width=3)
thickness_plot
dev.off()

thickness_plot
```



<!-- ### Unadjusted CLRs -->


```{r}
cordf.all <- cbind(info[, c("sample_barcode", "total.cells", "cell_density", "shannon_diversity")], t(cobj$E)) %>%
  inner_join(thicknessdf)
```


<!-- ```{r} -->
<!-- cordf <- cordf.all -->
<!-- rownames(cordf) <- cordf$sample_barcode -->
<!-- cordf$sample_barcode <- NULL -->
<!-- sample_table <- info %>% dplyr::select(donor_id, sample_barcode, anatomic_site) %>% unique() -->
<!-- prop.table <- cordf.all %>% inner_join(sample_table) -->

<!-- nosole <- prop.table[prop.table$anatomic_site != 'Sole',] -->
<!-- noscalpnosole <- prop.table[! prop.table$anatomic_site %in% c('Sole', 'Cen_Scalp', 'Occ_Scalp'),] -->

<!-- data_list <- list(ALL=prop.table, WITHOUT_SOLE=nosole, WITHOUT_SCALP_OR_SOLE=noscalpnosole) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- thickness.cols <- c("epidermis", "stratum.corneum", "total_thickness") -->
<!-- cell.cols <- cell_types -->


<!-- correlationdf <- lapply(names(data_list), function(condition){ -->
<!--   df <- data_list[[condition]] -->

<!--   resdf <- lapply(thickness.cols, function(thickness){ -->
<!--     # print(thickness) -->

<!--     spearmandf <- lapply(cell.cols, function(celltype){ -->
<!--      #  print(celltype) -->
<!--       res <- cor.test(df[, thickness], df[, celltype], method = 'spearman') %>% -->
<!--         broom::tidy() %>% -->
<!--         mutate(condition = condition, -->
<!--                cell_type.detailed = celltype, -->
<!--                measurement = thickness, -->
<!--                spearman.rho = estimate, -->
<!--                spearman.pval = p.value, -->
<!--                spearman.padj = p.adjust(p.value, method = "BH") -->
<!--                ) %>% -->
<!--         dplyr::select(condition, cell_type.detailed, measurement, spearman.rho, spearman.pval, spearman.padj) %>% -->
<!--         unique() %>% -->
<!--         data.frame() -->
<!--       return(res) -->
<!--     }) %>% bind_rows() -->

<!--       pearsondf <- lapply(cell.cols, function(celltype){ -->
<!--         # print(celltype) -->

<!--         res <- cor.test(df[, thickness], df[, celltype], method = 'pearson') %>% -->
<!--           broom::tidy() %>% -->
<!--           mutate(condition = condition, -->
<!--                  cell_type.detailed = celltype, -->
<!--                  measurement = thickness, -->
<!--                  pearson.r2 = estimate, -->
<!--                  pearson.pval = p.value, -->
<!--                  pearson.padj = p.adjust(p.value, method = "BH") -->
<!--           ) %>% -->
<!--           dplyr::select(condition, cell_type.detailed, measurement, pearson.r2, pearson.pval, pearson.padj) %>% -->
<!--           unique() %>% -->
<!--           data.frame() -->
<!--         return(res) -->
<!--       }) %>% -->
<!--         bind_rows() -->

<!--       res <- inner_join(spearmandf, pearsondf) -->
<!--       return(res) -->


<!--       }) %>% bind_rows() -->

<!--   return(resdf) -->

<!-- }) %>% bind_rows() %>% -->
<!--   mutate(tissue_compartment = 'dermis', -->
<!--          pearson.padj.star = gtools::stars.pval(pearson.padj), -->
<!--          spearman.padj.star = gtools::stars.pval(spearman.padj), -->
<!--          pearson.pval.star = gtools::stars.pval(pearson.pval), -->
<!--          spearman.pval.star = gtools::stars.pval(spearman.pval) -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r fig.height=8, fig.width=15} -->
<!-- corplot_thickness <- ((ggplot(correlationdf) + -->
<!--                       aes(x = measurement, y = cell_type.detailed, fill = spearman.rho, label = spearman.padj.star) + -->
<!--                       geom_tile() + geom_text() + -->
<!--                       theme_classic() + theme + rotate_x_text(angle=45) + -->
<!--                       scale_fill_gradientn(colors = c("blue", "white", 'red'), limits = c(-1, 1)) + -->
<!--                       facet_wrap(~factor(condition, levels = names(data_list)), nrow=1) + labs(title = "Spearman", x = NULL, y = NULL) ) | -->
<!--                      (ggplot(correlationdf) + -->
<!--                         aes(x = measurement, y = cell_type.detailed, fill = pearson.r2, label = pearson.padj.star) + -->
<!--                         geom_tile() + geom_text() + -->
<!--                         theme_classic() + theme +rotate_x_text(angle=45) + -->
<!--                         scale_fill_gradientn(colors = c("blue", "white", 'red'), limits = c(-1, 1)) + -->
<!--                         facet_wrap(~factor(condition, levels = names(data_list)), nrow=1) + labs(title = "Pearson", x = NULL, y = NULL))) + -->
<!--   plot_layout(guides='collect') -->

<!-- corplot_thickness -->
<!-- ``` -->


<!-- ```{r} -->
<!-- ggplot(cordf) + aes(x = `Papil Fib`, y = log1p(epidermis)) + geom_point() + geom_smooth(method='lm') + theme_bw() -->
<!-- ggplot(cordf) + aes(x = `Papil Fib`, y = log1p(total_thickness)) + geom_point() + geom_smooth(method='lm') + theme_bw() -->
<!-- ggplot(cordf) + aes(x = `Papil Fib`, y = log1p(epidermis)) + geom_point() + geom_smooth(method='lm') + theme_bw() -->
<!-- ggplot(cordf) + aes(x = `Papil Fib`, y = log1p(stratum.corneum)) + geom_point() + geom_smooth(method='lm') + theme_bw() -->
<!-- ``` -->


### Adjusted CLRs


```{r}
cordf.all <- cbind(info[, c("sample_barcode", "total.cells", "cell_density", "shannon_diversity")], adj_propmat2) %>%
  inner_join(thicknessdf) 
```


```{r}
cordf <- cordf.all
rownames(cordf) <- cordf$sample_barcode
cordf$sample_barcode <- NULL
sample_table <- info %>% dplyr::select(donor_id, sample_barcode, anatomic_site) %>% unique()
prop.table <- cordf.all %>% inner_join(sample_table)

nosole <- prop.table[prop.table$anatomic_site != 'Sole',]
noscalpnosole <- prop.table[! prop.table$anatomic_site %in% c('Sole', 'Cen_Scalp', 'Occ_Scalp'),]

data_list <- list(ALL=prop.table, WITHOUT_SOLE=nosole, WITHOUT_SCALP_OR_SOLE=noscalpnosole)
```


```{r}
thickness.cols <- c("epidermis", "stratum.corneum", "total_thickness")
cell.cols <- cell_types


correlationdf <- lapply(names(data_list), function(condition){
  df <- data_list[[condition]]
  
  resdf <- lapply(thickness.cols, function(thickness){
    # print(thickness)

    spearmandf <- lapply(cell.cols, function(celltype){
     #  print(celltype)
      res <- cor.test(log10(df[, thickness]), df[, celltype], method = 'spearman', exact=TRUE) %>%
        broom::tidy() %>%
        mutate(condition = condition,
               cell_type.detailed = celltype,
               measurement = thickness,
               spearman.rho = estimate,
               spearman.pval = p.value,
               spearman.padj = p.adjust(p.value, method = "BH")
               ) %>%
        dplyr::select(condition, cell_type.detailed, measurement, spearman.rho, spearman.pval, spearman.padj) %>%
        unique() %>%
        data.frame()
      return(res)
    }) %>% bind_rows()

      pearsondf <- lapply(cell.cols, function(celltype){
        # print(celltype)
        
        res <- cor.test(log10(df[, thickness]), df[, celltype], method = 'pearson', exact=TRUE) %>%
          broom::tidy() %>%
          mutate(condition = condition,
                 cell_type.detailed = celltype,
                 measurement = thickness,
                 pearson.r2 = estimate,
                 pearson.pval = p.value,
                 pearson.padj = p.adjust(p.value, method = "BH")
          ) %>%
          dplyr::select(condition, cell_type.detailed, measurement, pearson.r2, pearson.pval, pearson.padj) %>%
          unique() %>%
          data.frame()
        return(res)
      }) %>%
        bind_rows()
      
      res <- inner_join(spearmandf, pearsondf)
      return(res)


      }) %>% bind_rows()

  return(resdf)

}) %>% bind_rows() %>%
  mutate(tissue_compartment = 'dermis',
         pearson.padj.star = gtools::stars.pval(pearson.padj),
         spearman.padj.star = gtools::stars.pval(spearman.padj),
         pearson.pval.star = gtools::stars.pval(pearson.pval),
         spearman.pval.star = gtools::stars.pval(spearman.pval)
  )
```

```{r fig.height=8, fig.width=15}
corplot_thickness <- ((ggplot(correlationdf) +
                      aes(x = measurement, y = cell_type.detailed, fill = spearman.rho, label = spearman.padj.star) +
                      geom_tile() + geom_text() +
                      theme_classic() + theme + rotate_x_text(angle=45) +
                      scale_fill_gradientn(colors = c("blue", "white", 'red'), limits = c(-1, 1)) +
                      facet_wrap(~factor(condition, levels = names(data_list)), nrow=1) + labs(title = "Spearman", x = NULL, y = NULL) ) |
                     (ggplot(correlationdf) +
                        aes(x = measurement, y = cell_type.detailed, fill = pearson.r2, label = pearson.padj.star) +
                        geom_tile() + geom_text() +
                        theme_classic() + theme +rotate_x_text(angle=45) +
                        scale_fill_gradientn(colors = c("blue", "white", 'red'), limits = c(-1, 1)) +
                        facet_wrap(~factor(condition, levels = names(data_list)), nrow=1) + labs(title = "Pearson", x = NULL, y = NULL))) +
  plot_layout(guides='collect', axes='collect')

corplot_thickness
```


```{r}
ggplot(cordf) + aes(x = `Papil Fib`, y = log10(epidermis)) + geom_point() + geom_smooth(method='lm') + theme_bw()
ggplot(cordf) + aes(x = `Papil Fib`, y = log10(stratum.corneum)) + geom_point() + geom_smooth(method='lm') + theme_bw()
ggplot(cordf) + aes(x = `Papil Fib`, y = log10(total_thickness)) + geom_point() + geom_smooth(method='lm') + theme_bw()
```


```{r fig.height=5, fig.width=3}
epi <- ggplot(cordf) + 
  aes(x = `Papil Fib`, y = epidermis) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()

str <- ggplot(cordf) + 
  aes(x = `Papil Fib`, y = stratum.corneum) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()

papil <- (epi / str ) + plot_layout(axes = 'collect_x')

svglite::svglite('figures/merfish/BAYSOR/differential_abundance/FIG3F_PapilFib_EpidermalThicknessCorrelation.svg', height=5, width=3)
papil
dev.off()

papil
```


```{r fig.height=5, fig.width=3}
epi <- ggplot(cordf) + 
  aes(x = `Retic Fib II`, y = epidermis) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()

str <- ggplot(cordf) + 
  aes(x = `Retic Fib II`, y = stratum.corneum) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()



retic2 <- (epi / str ) + plot_layout(axes = 'collect_x')

svglite::svglite('figures/merfish/BAYSOR/differential_abundance/FIG3F_ReticFibII_EpidermalThicknessCorrelation.svg', height=5, width=3)
retic2
dev.off()

retic2
```

```{r fig.height=5, fig.width=3}
epi <- ggplot(cordf) + 
  aes(x = `Spn KC II`, y = epidermis) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()

str <- ggplot(cordf) + 
  aes(x = `Spn KC II`, y = stratum.corneum) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()

spn2 <- (epi / str ) + plot_layout(axes = 'collect_x')
svglite::svglite('figures/merfish/BAYSOR/differential_abundance/FIG3F_SpnKCII_EpidermalThicknessCorrelation.svg', height=5, width=3)
spn2
dev.off()

spn2
```

### Pairwise correlations of abundance


```{r}
library(corrplot)
```

```{r fig.height=10, fig.width=10}
cor.rmat <- cor(adj_propmat2, method='spearman')
cor.pmat <- cor.mtest(adj_propmat2, method='spearman')$p
cor.pstar <- gsub("\\.", "", gtools::stars.pval(cor.pmat))

## outline the cells that are significant (P < 0.05)
hm <- Heatmap(cor.rmat, name='Spearman',   
        cell_fun = function(j, i, x, y, width, height, fill) {
          if (cor.pmat[i, j] < 0.05){
            grid.rect(x = x, y = y, width = width, height = height, 
            gp = gpar(col = "black", fill = NA))
          }
        },
        width = ncol( as.matrix(cor.rmat))*unit(3.5, "mm"), 
        height = nrow( as.matrix(cor.rmat))*unit(3.5, "mm")
        )


svglite::svglite("figures/merfish/BAYSOR/differential_abundance/Fig3F_FINAL_AdjCLR_PairwiseCorrelatioN_Heatmap.svg", height=10, width=10)
draw(hm)
dev.off()

draw(hm)
```


## Per-Compartment Analysis


```{r make-per-compartment-objects}
sce <- SingleCellExperiment(assays = list(counts = obj[['RNA']]$counts), 
                            colData = obj@meta.data)

pb <- aggregateToPseudoBulk(sce,
                            assay = "counts",
                            cluster_id = "cell_type.detailed",
                            sample_id = "sample_compartment",
                            verbose = FALSE)
assayNames(pb) <- names(cell_types)
pb$sample_compartment <- colnames(pb)
counts <- cellCounts(pb)
colnames(counts) <- names(cell_types)
rm(sce); gc()

compartment_info <- read.csv("data/merfish/BAYSOR/metadata/tissue_compartment_areas.csv")
compartment_info$sample_barcode <- substr(compartment_info$sample_compartment, 1, nchar(compartment_info$sample_compartment)-4)

tx_metrics <- obj@meta.data %>% 
  group_by(sample_compartment) %>% 
  summarise(total_tx = sum(nCount_RNA), 
            mean_tx_per_cell = mean(nCount_RNA)) %>%
  ungroup()

info <- data.frame(colData(pb))
info$sample_compartment <- rownames(info)
info$total_area_mm2 <- NULL
info$sample_neighborhood <- NULL
info <- info %>% 
  inner_join(compartment_info) %>% 
  inner_join(tx_metrics) %>% 
  data.frame()
rownames(info) <- info$sample_compartment
info$anatomic_site <- make.names(info$anatomic_site)
counts <- counts[info$sample_compartment,]
```


### Direct Density and Diversity Boxplots

```{r create-density-matrix}
density <- t(sapply(rownames(info), function(x) {(counts[x,] / info[x, "compartment_area_mm2"])}))
```

```{r calculate-diveristy-index}
info$shannon_diversity <- diversity(t(density), index = 'shannon')$shannon
info$diveristy.density_adj <- info$shannon_diversity
info$total_cells <- rowSums(counts)
info$compartment_density <- (info$total_cells / (info$compartment_area_mm2)) / 100 # density is per 100um^2
```

```{r summarise-across-replicates-percompartment}
info2 <- info %>% 
  group_by(donor_id, compartment.short, tissue_compartment, site_short) %>% 
  summarise(shannon_diversity=mean(shannon_diversity), 
            compartment_density=mean(compartment_density),
            mean_tx_detected = mean(total_tx),
            mean_tx_per_cell = mean(mean_tx_per_cell))
```



```{r per-compartment-density-diversity-boxplots, fig.height=5, fig.width=6}
df <-info2[info2$compartment.short != 'STR', ]
df$compartment.short <- factor(df$compartment.short, levels = c("EPI", "DER", "SUB"))
df$site_short <- factor(df$site_short, levels = site_order)
div <- ggplot(df) + 
  aes(x = site_short, y = shannon_diversity, color = site_short) + 
  geom_quasirandom(show.legend=FALSE)+
  geom_boxplot(alpha=0, aes(color=NULL), show.legend=FALSE, outliers = FALSE) + 
  facet_wrap(~compartment.short, nrow=1) + 
  theme_classic() +
  scale_color_paletteer_d("Polychrome::light") +
  labs(x = NULL, y = "Shannon Diversity Index") + 
  theme + rotate_x_text()

den <- ggplot(df) + 
  aes(x = site_short, y = compartment_density, color = site_short) + 
  geom_quasirandom(show.legend=FALSE)+
  geom_boxplot(alpha=0, aes(color=NULL), show.legend=FALSE, outliers = FALSE) + 
  facet_wrap(~compartment.short, nrow=1) + 
  theme_classic() + 
  scale_color_paletteer_d("Polychrome::light") +
  labs(x = NULL, y = "Density (Cells/100um2)") + 
  theme + rotate_x_text() #+
  scale_y_log10()

FIG3D <- (div / den) + plot_layout(axes = 'collect_x') 
FIG3D
svglite::svglite("figures/merfish/BAYSOR/shannon_diversity/FIG3D_FINAL.tissue_area_normalized.by_compartments.shannon_index_and_density_panel.svg", height=5, width=6)
FIG3D
dev.off()
```

### Do Crumblr CLR Transformation


```{r}
cobj <- crumblr(counts)
```

#### Compartment CLR PCA


```{r}
pca <- prcomp(t(standardize(cobj)), scale. = TRUE, center=TRUE)
variance_proportions = round(summary(pca)$importance[2,]*100, digits=2)

# merge with metadata
df_pca <- merge(pca$x, info, by = "row.names")
```


```{r fig.height=6.5, fig.width=30}


# Plot PCA
anatomic_site.pca_plot <- ggplot(df_pca) +
  aes(x = PC1, y = PC2, 
      color = anatomic_site) +
  geom_point(size = 3) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) +
  scale_color_paletteer_d("Polychrome::light") 

batch.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2, color = batch, fill=batch, shape = gene_panel) +
  geom_point(stroke=2) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  scale_shape_manual(values = c(4,1)) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 

diversity.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2) +
  geom_density_2d(aes(color=compartment.short), contour_var = 'count', bins=50, size=0.1, alpha=0.5) +
  geom_point(aes(fill = shannon_diversity, size = compartment_density), shape=21) +
  viridis::scale_fill_viridis(option="A", name='Diversity') +
  scale_color_manual(values=c('SUB'="dodgerblue", 'DER'='goldenrod1', 'EPI'='orchid2'), 
                     name='Tissue\nCompartment') +
  scale_size_continuous(name='Density') +
  theme_classic() + theme + theme(aspect.ratio = 1,  legend.box='horizontal') +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 

compartment.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2, 
      color = compartment.short, 
      size = compartment_density) +
  geom_point() +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 

(anatomic_site.pca_plot |
  batch.pca_plot |
  diversity.pca_plot |
  compartment.pca_plot ) & theme

```

```{r}
diversity.pca_plot <- ggplot(df_pca) +
  aes(PC1, PC2) +
  geom_density_2d(aes(color=compartment.short), contour_var = 'count', bins=50, size=0.1, alpha=0.5) +
  geom_point(aes(fill = shannon_diversity, size = compartment_density), shape=21) +
  viridis::scale_fill_viridis(option="A", name='Diversity') +
  scale_color_manual(values=c('SUB'="dodgerblue", 'DER'='goldenrod1', 'EPI'='orchid2'), 
                     name='Tissue\nCompartment') +
  scale_size_continuous(name='Density') +
  theme_classic() + theme + theme(aspect.ratio = 1,  legend.box='horizontal') +
  labs(x = paste0("PC1: ", variance_proportions[1], "%"),
       y = paste0("PC2: ", variance_proportions[2], "%")) 

svglite::svglite("figures/merfish/BAYSOR/shannon_diversity/FIG3E.FINAL.per_compartment.density_diversity.crumblr_clr.pca_plot.svg", height=5, width=7)
diversity.pca_plot
dev.off()

diversity.pca_plot
```


## Epidermal thickness vs dermal proportions


```{r}
cordf.all <- data.frame(cbind(info[, c("sample_barcode", 'compartment.short')], t(cobj$E))) %>%
  filter(compartment.short =='DER') %>%
  inner_join(thicknessdf) 
```


```{r}
cordf <- cordf.all
rownames(cordf) <- cordf$sample_barcode
cordf$sample_barcode <- NULL
sample_table <- info %>% dplyr::select(donor_id, sample_barcode, anatomic_site) %>% unique()
prop.table <- cordf.all %>% inner_join(sample_table)

nosole <- prop.table[prop.table$anatomic_site != 'Sole',]
noscalpnosole <- prop.table[! prop.table$anatomic_site %in% c('Sole', 'Cen_Scalp', 'Occ_Scalp'),]

data_list <- list(ALL=prop.table, WITHOUT_SOLE=nosole, WITHOUT_SCALP_OR_SOLE=noscalpnosole)
```


```{r}
thickness.cols <- c("epidermis", "stratum.corneum", "total_thickness")
cell.cols <- names(cell_types)


correlationdf <- lapply(names(data_list), function(condition){
  df <- data_list[[condition]]
  
  resdf <- lapply(thickness.cols, function(thickness){
    # print(thickness)

    spearmandf <- lapply(cell.cols, function(celltype){
     #  print(celltype)
      res <- cor.test(df[, thickness], df[, celltype], method = 'spearman') %>%
        broom::tidy() %>%
        mutate(condition = condition,
               cell_type.detailed = celltype,
               measurement = thickness,
               spearman.rho = estimate,
               spearman.pval = p.value,
               spearman.padj = p.adjust(p.value, method = "BH")
               ) %>%
        dplyr::select(condition, cell_type.detailed, measurement, spearman.rho, spearman.pval, spearman.padj) %>%
        unique() %>%
        data.frame()
      return(res)
    }) %>% bind_rows()

      pearsondf <- lapply(cell.cols, function(celltype){
        # print(celltype)
        
        res <- cor.test(df[, thickness], df[, celltype], method = 'pearson') %>%
          broom::tidy() %>%
          mutate(condition = condition,
                 cell_type.detailed = celltype,
                 measurement = thickness,
                 pearson.r2 = estimate,
                 pearson.pval = p.value,
                 pearson.padj = p.adjust(p.value, method = "BH")
          ) %>%
          dplyr::select(condition, cell_type.detailed, measurement, pearson.r2, pearson.pval, pearson.padj) %>%
          unique() %>%
          data.frame()
        return(res)
      }) %>%
        bind_rows()
      
      res <- inner_join(spearmandf, pearsondf)
      return(res)


      }) %>% bind_rows()

  return(resdf)

}) %>% bind_rows() %>%
  mutate(tissue_compartment = 'dermis',
         pearson.padj.star = gtools::stars.pval(pearson.padj),
         spearman.padj.star = gtools::stars.pval(spearman.padj),
         pearson.pval.star = gtools::stars.pval(pearson.pval),
         spearman.pval.star = gtools::stars.pval(spearman.pval)
  )
```

```{r fig.height=8, fig.width=15}
corplot_thickness <- ((ggplot(correlationdf) +
                      aes(x = measurement, y = cell_type.detailed, fill = spearman.rho, label = spearman.padj.star) +
                      geom_tile() + geom_text() +
                      theme_classic() + theme + rotate_x_text(angle=45) +
                      scale_fill_gradientn(colors = c("blue", "white", 'red'), limits = c(-1, 1)) +
                      facet_wrap(~factor(condition, levels = names(data_list)), nrow=1) + labs(title = "Spearman", x = NULL, y = NULL) ) |
                     (ggplot(correlationdf) +
                        aes(x = measurement, y = cell_type.detailed, fill = pearson.r2, label = pearson.padj.star) +
                        geom_tile() + geom_text() +
                        theme_classic() + theme +rotate_x_text(angle=45) +
                        scale_fill_gradientn(colors = c("blue", "white", 'red'), limits = c(-1, 1)) +
                        facet_wrap(~factor(condition, levels = names(data_list)), nrow=1) + labs(title = "Pearson", x = NULL, y = NULL))) +
  plot_layout(guides='collect')

corplot_thickness
```


```{r fig.height=3, fig.width=6}
epi <- ggplot(cordf) + 
  aes(x = Papil.Fib, y = epidermis) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()

str <- ggplot(cordf) + 
  aes(x = Papil.Fib, y = stratum.corneum) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_y_log10()


svglite::svglite('figures/merfish/BAYSOR/differential_abundance/FIG3F_PapilFib_EpidermalThicknessCorrelation.svg', height=5, width=3)
(epi / str ) + plot_layout(axes = 'collect_x')
dev.off()
```

```{r fig.height=3, fig.width=8}
epi <- ggplot(cordf) + 
  aes(y = Papil.Fib, x = epidermis) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_x_log10()

str <- ggplot(cordf) + 
  aes(y = Papil.Fib, x = stratum.corneum) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme +
  scale_x_log10()

tot <- ggplot(cordf) + 
  aes(y = Papil.Fib, x = total_thickness) + 
  geom_point() + 
  geom_smooth(data=cordf, color='red', method='lm', se=FALSE) + # all samples
  geom_smooth(data=nosole, color='blue', method='lm', se=FALSE) +
  theme_bw() + theme  +
  scale_x_log10()


(epi | str | tot) + plot_layout(axes = 'collect_y')
```

